---
title: "BS_2"
author: "fionarhuang"
date: "2020-06-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
    library(ggplot2)
    library(gganimate)
    library(ggtree)
    library(dplyr)
    library(treeclimbR)  
    library(ape)
    library(TreeHeatmap)
    library(ggnewscale)
    library(cowplot)
    library(phylofactor)
  library(treeSeg)
})
```


# Data simulation

We simulate a scenario (`BS`) that only two branches on the tree have
differential abundance (signal) between groups. The data is simulated by
swapping proportions of two branches, and leaves in the same branch have the
same fold change.

```{r}
# generate a random tree
set.seed(1)
n <- 100
tree <- rtree(n)


# generate a random probility vector  for leaves
p <- rbeta(n = n, shape1 = 2, shape2 = 5)
p <- p/sum(p)
names(p) <- tree$tip.label

# # simulate counts by sampling from multinomial distribution
lse <- simMult(pr = p, libSize = 1000, tree = tree,
               minTip.A = 10, maxTip.A = 20,
               ratio = 2, scenario = "BS",
               nSam = c(10, 10))

```

## Viz simulated data

DA branches are colored in orange or blue. Simulated counts are scaled and displayed in the heatmap
```{r}
# signal branches
br <- metadata(lse)$branch
source("code/viewSim.R")
treeFig <- viewSim(lse, branch.length = "none",
                   zoom_scale = 1, size = 1.1)

# scale counts
count <- assays(lse)[[1]]
rownames(count) <- rowLinks(lse)$nodeLab

scale_count <- t(apply(count, 1, FUN = function(x) {
    xx <- scale(x)
    (xx - min(xx))/(max(xx)-min(xx))
}))
rownames(scale_count) <- rownames(count)
colnames(scale_count) <- colnames(count)


# fig: tree + heatmap
vv <- gsub(pattern = "_.*", "", colnames(count))
names(vv) <- colnames(scale_count)
fig <- TreeHeatmap(tree = tree, tree_fig = treeFig, 
                   hm_data = scale_count,
                   column_split = vv, rel_width = 0.7, 
                   tree_hm_gap = 0.3) +
    scale_fill_viridis_c(option = "B") +
  theme(legend.position = c(0.1, 0.8))
fig
```


# `PhyloFactor`

```{r,fig.height=8}
site <- colData(lse)$group
```

Different number of factors are used.
```{r phylofactor, cache=TRUE, message=FALSE}
pf_5 <- PhyloFactor(count, tree, site, nfactors= 5)
f_5 <- pf.tree(pf_5, tree = tree, layout='rectangular', 
               branch.length = "none",
               ignore.tips = FALSE)$ggplot


pf_10 <- PhyloFactor(count, tree, site, nfactors= 10)
f_10 <- pf.tree(pf_10, tree = tree,layout='rectangular', 
                branch.length = "none",
                ignore.tips = FALSE)$ggplot

pf_50 <- PhyloFactor(count, tree, site, nfactors= 50)
f_50 <- pf.tree(pf_10, tree = tree,layout='rectangular', 
                branch.length = "none",
               ignore.tips = FALSE)$ggplot

pf_NULL <- PhyloFactor(count, tree, site, nfactors= NULL)
f_NULL <- pf.tree(pf_NULL, tree = tree,layout='rectangular',
                  branch.length = "none",
                  ignore.tips = FALSE)$ggplot

# summary(pf_PhyloFactor,factor=1)
# pf_PhyloFactor$factors
```

## Truth vs phylofactor
```{r,fig.height=6, fig.width=12}
plot_grid(treeFig, f_5, f_10, f_50, f_NULL, 
          nrow = 1, 
          labels = c("Truth", "nf = 5", "nf = 10", "nf = 50", "nf = NULL"), 
          label_size = 7)

```


# treeclimbR

## data aggregation
```{r}
all_node <- showNode(tree = rowTree(lse), only.leaf = FALSE)
tse <- aggValue(x = lse, rowLevel = all_node)
colData(tse)
```

## differential analysis

```{r}
res <- runDA(TSE = tse, feature_on_row = TRUE, 
             filter_min_count = 0, 
             design_terms = "group", normalize = FALSE)
out <- nodeResult(object = res, n = Inf)
head(out)
dim(out)
```

## run treeclimbR
```{r treeclimbR}
# treeclimbR
cand <- getCand(tree = rowTree(tse), score_data = out, 
                node_column = "node", p_column = "PValue",
                sign_column = "logFC", message = TRUE)

best <- evalCand(tree = rowTree(tse), levels = cand$candidate_list, 
                 score_data = out, node_column = "node",
                 sign_column = "logFC",
                 p_column = "PValue")
infoCand(best)


```


## result

```{r}
# the detected nodes
loc <- best$output[best$output$signal.node, ][["node"]]
loc

fig_climb <- treeFig +
  geom_point2(aes(subset = (node %in% loc)), color = "red", size = 2) +
  theme(legend.position = "none")
```

# treeSeg

As `treeSeg` requires to specify the distribution of data, and accepts only `gauss` and `binomial`, we here transform P-value to Z-score to use `gauss`.

```{r treeSeg, cache=TRUE}
out_leaf <- out %>%
  filter(node %in% 1:100) %>%
  arrange(node)

# check order
all(out_leaf$node == 1:100)

pval <- out_leaf$PValue
tp <- 1-pval
# 10E-8 is to avoid Inf value in z that treeSeg doesn't accept
tp <- ifelse(tp == 0, 10E-8, ifelse(tp == 1, 1-10E-8, tp))
z <- qnorm(tp)
names(z) <- transNode(tree = tree, node = out_leaf$node)

ansGauss <- treeSeg(z, tree, alpha = 0.05, fam = "gauss")
ansGauss$mlAN

fig_seg <- treeFig +
  geom_point2(aes(subset = node %in% ansGauss$mlAN), 
              color = "red")
```


# Truth vs methods
```{r,fig.height=7, fig.width=12}
plot_grid(fig, 
          fig_seg + theme(legend.position = "none"), 
          fig_climb, f_5, f_10, f_NULL, 
          nrow = 1, rel_widths = c(0.28, rep(0.145, 5)),
          labels = c("", "treeSeg", "treeclimbR", "nf = phyfact_5",
                      "nf = phyfact_10", "nf = phyfact_NULL"), 
          label_size = 9)

```

